A Theory of Bidirectional Type Synthesis for Simple Types

### 2023-06-14 tex/sec3-defs.tex
[ tex ] Fix the colour of modes; use the verb form of decidable.
For example, instead of saying “for every type A it is decidable whether …”, we can say “every type A decides whether …”
[ tex ] Fix typos and remove old requirements

  For brevity, we write $o \colon \Xi \rhd [\Delta_1]A_{1}^{d_1}, \ldots, [\Delta_{n}] A^{d_n}_{n} \to A^{d}$ to indicate an operation with its arity. 
\end{definition}

@@

\begin{figure}
  \centering
  \small
  \judgbox{|-_{\Sigma, \Omega} \isTerm{t}^\dir{d}}{$t$ is a raw term for $(\Sigma, \erase{\Omega})$ in mode $d$ for  $(\Sigma, \Omega)$}
  \begin{mathpar}
    \inferrule{x : \Identifier}{|-_{\Sigma, \Omega} \isTerm{x}^{\syn}}\,\SynRule{Var}
    \and
    \inferrule{\cdot |-_{\Sigma} A \\ |-_{\Sigma, \Omega}\isTerm{t}^{\chk}}{|-_{\Sigma, \Omega} (\isTerm{t \annote A})^{\syn}}\,\SynRule{Anno}
    \and
    \inferrule{|-_{\Sigma, \Omega} \isTerm{t}^{\syn}}{|-_{\Sigma, \Omega} \isTerm{t}^{\chk}}\,\ChkRule{Sub}
  \end{mathpar}
  \begin{mathpar}
    \inferrule{|-_{\Sigma, \Omega} \isTerm{t_1}^\dir{d_1} \quad \cdots \quad |-_{\Sigma, \Omega} \isTerm{t_n}^\dir{d_n}}
    {|-_{\Sigma, \Omega} \isTerm{\tmOp_o(\vec{x}_1.\, t_1; \ldots;\vec{x}_n.\, t_n)}^\dir{d}}\,\Rule{Op}
    \and \text{for $o \colon \Xi \rhd [\Delta_1]A_{1}^\dir{d_1}, \ldots, [\Delta_{n}] A_{n}^\dir{d_n} \to A^\dir{d}$ in $\Omega$}
  \end{mathpar}
  \caption{Preprocessed Raw terms}
\end{figure}

\begin{figure}
  \centering
  \small
  \judgbox{\Gamma |-_{\Sigma, \Omega} \isTerm{t} :^\dir{d} A}{A raw term $t$ in mode $d$ has a type $A$ under $\Gamma$}
  \begin{mathpar}
    \inferrule{(x : A) \in \Gamma}{\Gamma |-_{\Sigma, \Omega} \isTerm{x} :^{\syn} A}\,\SynRule{Var}
    \and
    \inferrule{\Gamma |-_{\Sigma, \Omega} \isTerm{t} :^{\chk} A}{\Gamma |-_{\Sigma, \Omega} (\isTerm{t \annote B}) :^{\syn}  A}\,\SynRule{Anno}
    \and
    \inferrule{\Gamma |-_{\Sigma, \Omega} \isTerm{t} :^{{\syn}} B \\ A = B}{\Gamma |-_{\Sigma, \Omega} \isTerm{t} :^{\chk} A}\,\ChkRule{Sub}
    \\
    \inferrule{\rho\colon \Sub{\Xi}{\emptyset} \\ \Gamma, \simsub{\isTerm{\vec{x}_{1}} : \Delta_{1}}{\rho} |-_{\Sigma, \Omega} \isTerm{t_{1}} :^\dir{d_1} \simsub{A_1}{\rho} \\
      \cdots \\
    \Gamma, \simsub{\isTerm{\vec{x}_{n}} : \Delta_{n}}{\rho} |-_{\Sigma, \Omega} \isTerm{t_n} :^{\dir{d_n}} \simsub{A_{n}}{\rho}}
    {\Gamma |-_{\Sigma, \Omega} \isTerm{\tmOp_o(\vec{x}_1 .\, t_{1}; \ldots; \vec{x}_{n}.\, t_{n})} :^{\dir{d}} \simsub{A}{\rho}} \,\mathsf{Op}
    \and \text{for $o \colon \Xi \rhd [\Delta_1]A_{1}^\dir{d_1}, \ldots, [\Delta_{n}] A_{n}^\dir{d_n} \to A^\dir{d}$ in $\Omega$}
  \end{mathpar}
  \begin{minipage}{.15\linewidth}
  Abbreviations: 
  \end{minipage}
  \begin{minipage}{.8\linewidth}
    \begin{flushleft}
    $\boxed{\Gamma |-_{\Sigma, \Omega} \isTerm{t} \syn A} \defeq \boxed{\Gamma |-_{\Sigma, \Omega} \isTerm{t} :^{\syn} A}$ \quad A raw term $t$ synthesises a type $A$ under $\Gamma$ \\
    $\boxed{\Gamma |-_{\Sigma, \Omega} \isTerm{t} \chk A} \defeq \boxed{\Gamma |-_{\Sigma, \Omega} \isTerm{t} :^{\chk} A}$ \quad A raw term $t$ checks against a type $A$ under $\Gamma$
    \end{flushleft}
  \end{minipage}
  \caption{Bidirectional typing rules}
\end{figure}

>>>>

  For brevity, we write $o \colon \Xi \rhd [\Delta_1]A_{1}^{\dir{d_1}}, \ldots, [\Delta_{n}] A^{\dir{d_n}}_{n} \to A^{\dir{d}}$ to indicate an operation with its arity. 
\end{definition}

@@

\begin{figure}
  \centering
  \small
  \judgbox{|-_{\Sigma, \Omega} \isTerm{t}^\dir{d}}{$t$ is a raw term for $(\Sigma, \erase{\Omega})$ in mode $d$ for  $(\Sigma, \Omega)$}
  \begin{mathpar}
    \inferrule{x : \Identifier}{|-_{\Sigma, \Omega} \isTerm{x}^{\syn}}\,\SynRule{Var}
    \and
    \inferrule{\cdot |-_{\Sigma} A \\ |-_{\Sigma, \Omega}\isTerm{t}^{\chk}}{|-_{\Sigma, \Omega} (\isTerm{t \annote A})^{\syn}}\,\SynRule{Anno}
    \and
    \inferrule{|-_{\Sigma, \Omega} \isTerm{t}^{\syn}}{|-_{\Sigma, \Omega} \isTerm{t}^{\chk}}\,\ChkRule{Sub}
  \end{mathpar}
  \begin{mathpar}
    \inferrule{|-_{\Sigma, \Omega} \isTerm{t_1}^\dir{d_1} \quad \cdots \quad |-_{\Sigma, \Omega} \isTerm{t_n}^\dir{d_n}}
    {|-_{\Sigma, \Omega} \isTerm{\tmOp_o(\vec{x}_1.\, t_1; \ldots;\vec{x}_n.\, t_n)}^\dir{d}}\,\Rule{Op}
    \and \text{for $o \colon \Xi \rhd [\Delta_1]A_{1}^{\dir{d_1}}, \ldots, [\Delta_{n}] A_{n}^{\dir{d_n}} \to A^{\dir{d}}$ in $\Omega$}
  \end{mathpar}
  \caption{Preprocessed raw terms}
\end{figure}

\begin{figure}
  \centering
  \small
  \judgbox{\Gamma |-_{\Sigma, \Omega} \isTerm{t} :^\dir{d} A}{A raw term $t$ has a type $A$ (without type variables) under $\Gamma$}
  \begin{mathpar}
    \inferrule{(x : A) \in \Gamma}{\Gamma |-_{\Sigma, \Omega} \isTerm{x} :^{\syn} A}\,\SynRule{Var}
    \and
    \inferrule{\Gamma |-_{\Sigma, \Omega} \isTerm{t} :^{\chk} A}{\Gamma |-_{\Sigma, \Omega} (\isTerm{t \annote B}) :^{\syn}  A}\,\SynRule{Anno}
    \and
    \inferrule{\Gamma |-_{\Sigma, \Omega} \isTerm{t} :^{{\syn}} B \\ A = B}{\Gamma |-_{\Sigma, \Omega} \isTerm{t} :^{\chk} A}\,\ChkRule{Sub}
    \\
    \inferrule{\rho\colon \Sub{\Xi}{\emptyset} \\ \Gamma, \simsub{\isTerm{\vec{x}_{1}} : \Delta_{1}}{\rho} |-_{\Sigma, \Omega} \isTerm{t_{1}} :^\dir{d_1} \simsub{A_1}{\rho} \\
      \cdots \\
    \Gamma, \simsub{\isTerm{\vec{x}_{n}} : \Delta_{n}}{\rho} |-_{\Sigma, \Omega} \isTerm{t_n} :^{\dir{d_n}} \simsub{A_{n}}{\rho}}
    {\Gamma |-_{\Sigma, \Omega} \isTerm{\tmOp_o(\vec{x}_1 .\, t_{1}; \ldots; \vec{x}_{n}.\, t_{n})} :^{\dir{d}} \simsub{A}{\rho}} \,\mathsf{Op}
    \and \text{for $o \colon \Xi \rhd [\Delta_1]A_{1}^{\dir{d_1}}, \ldots, [\Delta_{n}] A_{n}^{\dir{d_n}} \to A^\dir{d}$ in $\Omega$}
  \end{mathpar}
  \begin{minipage}{.15\linewidth}
  Abbreviations: 
  \end{minipage}
  \begin{minipage}{.8\linewidth}
    \begin{flushleft}
    $\boxed{\Gamma |-_{\Sigma, \Omega} \isTerm{t} \syn A} \defeq \boxed{\Gamma |-_{\Sigma, \Omega} \isTerm{t} :^{\syn} A}$ \quad A raw term $t$ synthesises a type $A$ under $\Gamma$ \\
    $\boxed{\Gamma |-_{\Sigma, \Omega} \isTerm{t} \chk A} \defeq \boxed{\Gamma |-_{\Sigma, \Omega} \isTerm{t} :^{\chk} A}$ \quad A raw term $t$ checks against a type $A$ under $\Gamma$
    \end{flushleft}
  \end{minipage}
  \caption{Bidirectional typing rules}
\end{figure}

### 2023-06-14 tex/sec5-synthesis.tex
[ tex ] Fix typos and remove old requirements

\begin{theorem}[Decidability of Bidirectional Typing Synthesis] \label{thm:bidirectional-type-synthesis-checking}
  Let $(\Sigma, \Omega)$ be a mode-correct bidirectional type system.
  For every context $\Gamma$ and raw term $\isTerm{t}$ in mode $d$, 
  \begin{enumerate}
    \item if $\isTerm{t}$ is synthesising, then it is decidable whether there is a type~$A$ and a derivation of
      \[
        \Gamma |-_{\Sigma, \Omega} \isTerm{t} \syn A.
      \]
    \item if $\isTerm{t}$ is checking, then every type $A$ decides whether there is a derivation of
      \[
        \Gamma |-_{\Sigma, \Omega} \isTerm{t} \chk A.
      \]
  \end{enumerate}
\end{theorem}
\begin{lemma}\label{lem:args-induction}
  Assume \Cref{thm:bidirectional-type-synthesis}.
  Then, for every set $\Xi$ of type variables, list 
  \[
    \mathit{as} = \left([\Delta_i] A_{i}^{\dir{d_i}}\right)_{i = 1}^{n}
  \]
  list of raw terms $t_i$'s in mode $d_i$, context $\Gamma$, and \emph{partial} substitution $\rho$ from $\Xi$ to $\emptyset$
  \begin{enumerate}
    \item either there is a minimal extension $\ext{\rho}$ of $\rho$ defined on $\synvar(\mathit{as})$ such that all of following judgements are derivable
      \[
        \Gamma, \simsub{\Delta_i}{\ext{\rho}} |-_{\Sigma, \Omega} \isTerm{t_i} \colon \simsub{A_i}{\ext{\rho}}^{\dir{d_i}}
      \]

>>>>

\begin{theorem}[Decidability of Bidirectional Type Synthesis and Checking] \label{thm:bidirectional-type-synthesis-checking}
  Let $(\Sigma, \Omega)$ be a mode-correct bidirectional type system.
  For every context $\Gamma$ and raw term $\isTerm{t}$ in mode $d$, 
  \begin{enumerate}
    \item if $\isTerm{t}$ is synthesising, then it is decidable whether there is a type~$A$ and a derivation of
      \[
        \Gamma |-_{\Sigma, \Omega} \isTerm{t} \syn A.
      \]
    \item if $\isTerm{t}$ is checking, then every type $A$ decides whether there is a derivation of
      \[
        \Gamma |-_{\Sigma, \Omega} \isTerm{t} \chk A.
      \]
  \end{enumerate}
\end{theorem}
\begin{lemma}\label{lem:args-induction}
  Assume \Cref{thm:bidirectional-type-synthesis}.
  Then, for every set $\Xi$ of type variables, list 
  \[
    \mathit{as} = \left([\Delta_i] A_{i}^{\dir{d_i}}\right)_{i = 1}^{n}
  \]
  list of raw terms $t_i$'s in mode $d_i$, context $\Gamma$, and \emph{partial} substitution $\rho$ from $\Xi$ to $\emptyset$
  \begin{enumerate}
    \item either there is a minimal extension $\ext{\rho}$ of $\rho$ defined on $\synvar(\mathit{as})$ such that all of following judgements are derivable
      \[
        \Gamma, \simsub{\Delta_i}{\ext{\rho}} |-_{\Sigma, \Omega} \isTerm{t_i} \colon \simsub{A_i}{\ext{\rho}}^{\dir{d_i}}
      \]
